# Components of time series and naive models

## What is time series?

A time series is a sequence of observations or random variables ordered by time.

## Time Series Components

-   Trend, cyclicity, seasonality
-   Additive and multiplicative decomposition

The **additive** decomposition of a time series is represented as:

$$Y_t = T_t + S_t + R_t$$

where:

- $Y_t$ is the observed value at time $t$,\
- $T_t$ is the trend component,\
- $S_t$ is the seasonal component,\
- $R_t$ is the residual (or random) component.

**Trend** is a smoothly changing component of a time series.

**Seasonality** is a strong periodic component with stable intensity.

The **multiplicative** decomposition of a time series is represented as:

$$Y_t = T_t \times S_t \times R_t$$

where:\
- $Y_t$ is the observed value at time $t$,\
- $T_t$ is the trend component,\
- $S_t$ is the seasonal component,\
- $R_t$ is the residual (or random) component.

Transform additive to multiplicative:

$$ln(Y_t) = ln(T_t \times S_t \times R_t) = ln(T_t) + ln(S_t) + ln(R_t)$$

## Algorithm STL

STL - Seasonal Trend Decomposition with Loess.

Loess - local linear regression.

## Time Series Characteristics

### ACF {-}

$\hat{y}_t = \hat{\beta}_1 + \hat{\beta}_2 y_{t-1}, ACF_1 = \hat{\beta}_2$

$\hat{y}_t = \hat{\beta}_1 + \hat{\beta}_2 y_{t-1}, ACF_2 = \hat{\beta}_2$

$\hat{y}_t = \hat{\beta}_1 + \hat{\beta}_2 y_{t-k}, ACF_k = \hat{\beta}_2$

$ACF_k$ is a sample correlation between $y_t$ and $y_{t-k}$

### PACF {-}

$\hat{y}_t = \hat{\alpha} + \hat{\beta}_1y_{t-1}, PACF_1 = \hat{\beta}_1$

$\hat{y}_t = \hat{\alpha} + \hat{\beta}_1 y_{t-1} + \hat{\beta}_2 y_{t-2}, PACF_2 = \hat{\beta}_2$

$\hat{y}_t = \hat{\alpha} + \hat{\beta}_1 y_{t-1} + ... + \hat{\beta}_k y_{t-k}, PACF_k = \hat{\beta}_k$

## Simple Models

### White Noise {-}

Time series $u_t$ is a white noise, if:

-   $\mathbb{E}(u_t) = 0$
-   $Var(u_t) = \sigma^2$
-   $Cov(u_s, u_t) = 0, s \neq t$

> White noise is a part of **each** statistical model.

### Model with independent observations {-}

$y_t = \mu + u_t$, where $u_t$ - white noise, $u_t \sim \mathcal{N}(0, \sigma^2)$

### Naive Model (Random Walk) {-}

$y_t = y_{t-1} + u_t$, where $u_t$ - white noise, $u_t \sim \mathcal{N}(0, \sigma^2)$, $y_1$ is known.

### Seasonal Random Walk {-}

$y_t = y_{t-12} + u_t$, where $u_t$ - white noise, $u_t \sim \mathcal{N}(0, \sigma^2)$, $y_1, ..., y_{11}$ are known.

## Examples

### Time series from scratch

Load libraries.

```{r 01-time-series-components-1 }
# loading libraries
library(tidyverse)
library(fpp3)

```

Create time series from scratch.

```{r 01-time-series-components-2 }

set.seed(777)
n_obs <- 120
two <- tsibble(date = yearmonth(ymd('2010-01-01') + months(0: (n_obs - 1))),
               iid = rnorm(n_obs, mean = 10, sd = 4),
               rwalk = 10 + cumsum(rnorm(n_obs, mean = 0, sd = 1)),
               index = date)
two

```

Have a look at the data.

```{r 01-time-series-components-3 }

two |>
  autoplot(iid) +
  xlab('Date') +
  ylab('IID') +
  ggtitle('Independent observations')

```

```{r 01-time-series-components-4 }

two |>
  autoplot(rwalk) +
  xlab('Date') +
  ylab('rwalk') +
  ggtitle('Random walk')

```

Seasonal plots.

```{r 01-time-series-components-5 }

two |> 
  gg_season(rwalk)

```

```{r 01-time-series-components-6 }

two |>
  gg_subseries(rwalk)

```

```{r 01-time-series-components-7 }

# random walk
two |>
  gg_lag(rwalk)

```

```{r 01-time-series-components-8 }

# independent observations
two |>
  gg_lag(iid)

```

```{r 01-time-series-components-9 }

two |> gg_tsdisplay(rwalk, plot_type = 'season')

```

### Time Series data cleaning

1. Libraries loading
```{r 01-time-series-components-10 }

# loading libs
library(tidyverse)
library(fpp3)
library(rio)

```
2. Data Loading

```{r 01-time-series-components-11 }

d <- import(paste0(getwd(), '/00_data/rf_marriage.xlsx'), skip = 2)
head(d)
glimpse(d)

```

3. Data cleaning

```{r 01-time-series-components-12 }

colnames(d)[1:3] <- c('region', 'unit', 'period')
glimpse(d)

```

```{r 01-time-series-components-13 }

d$period |> unique()

```
```{r 01-time-series-components-14 }

nchar(d$period |> unique())

```

```{r 01-time-series-components-15 }

d2 <- d |> filter(nchar(period) < 13)
d2$period |> unique()

```

```{r 01-time-series-components-16 }

month_dict <- tibble(period = unique(d2$period),
                     month = 1:12)
month_dict

```

```{r 01-time-series-components-17 }

d3 <- d2 |> left_join(month_dict, by = 'period')
head(d3)

```

```{r 01-time-series-components-18 }

d4 <- d3 |> select(-unit, -period)
head(d4)

```

```{r 01-time-series-components-19 }

d5 <- d4 |>
  pivot_longer(cols = `2006`:`2024`, names_to = 'year', values_to = 'total')
head(d5)
glimpse(d5)

```

```{r 01-time-series-components-20 }

d6 <- d5 |>
  mutate(date = yearmonth(paste0(year, '-', month)))
glimpse(d6)

```
```{r 01-time-series-components-21 }

d7 <- d6 |>
  separate(region, into = c('code', 'name'), sep = ' ', extra = 'merge')

glimpse(d7)

```

```{r 01-time-series-components-22 }

d8 <- d7 |>
  select(-month, - year)
glimpse(d8)

```
```{r 01-time-series-components-23 }

marriages <- as_tsibble(d8, index = date, key = c('code', 'name'))
marriages

```

```{r 01-time-series-components-24 }

# plot the data
marriages |>
  filter(code == 643) |> autoplot()

```
```{r 01-time-series-components-25 }

marriages |>
  mutate(date = as.Date(date)) |>
  export(paste0(getwd(), '/00_data/marriages.csv'))

```

```{r 01-time-series-components-26 }

m <- import(paste0(getwd(), '/00_data/marriages.csv'))
glimpse(m)

```
```{r 01-time-series-components-27 }

ts_marriages <- m |>
                mutate(date = yearmonth(date)) |>
                as_tsibble(index = date, key = c('code', 'name'))
glimpse(ts_marriages)

head(ts_marriages)

```

```{r 01-time-series-components-28 }

ts_marriages |>
  filter(code == 643) |>
  gg_tsdisplay(total, plot_type = 'season')

```

```{r 01-time-series-components-29 }

m_rf <- ts_marriages |>
  filter(code == 643)

head(m_rf)
tail(m_rf, 10)

```

```{r 01-time-series-components-30 }

m_rf_full <- m_rf |>
  filter(date < ymd('2024-06-01'))

m_rf_full |>
  gg_tsdisplay(total, plot_type = 'season')

```
4. Time Series Analysis

```{r 01-time-series-components-31 }

stl_model <- m_rf_full |>
  model(decomp = STL(total ~ trend(window = 7) + season(window = 100)))

components(stl_model)

```

```{r 01-time-series-components-32 }

stl_model |>
  components() |>
    autoplot()

```

```{r 01-time-series-components-33 }

m_rf_full |>
  model(decomp = STL(total ~ trend(window = 100) + season(window = 100))) |>
  components() |>
    autoplot()

```

```{r 01-time-series-components-34 }

m_rf_full |>
  model(decomp = STL(total ~ trend(window = 100) + season(window = 5))) |>
  components() |>
    autoplot()

```

5. Time series features

```{r 01-time-series-components-35 }

marr_features <- ts_marriages |>
  features(total, features = feat_stl)
           
glimpse(marr_features)

```

```{r 01-time-series-components-36, warning=FALSE}

library(ggrepel)

ggplot(marr_features, aes(x = trend_strength, y = seasonal_strength_year, label=name)) +
  geom_point() +
  geom_text_repel()
  

```

