# Exponential smoothing models

```{r 02-exponential-smoothing-1 }

library(tidyverse)
library(fpp3)
library(ggplot2)
library(rio)

```

```{r 02-exponential-smoothing-2 }

m <- import(paste0(getwd(), '/00_data/marriages.csv'))
glimpse(m)

```

```{r 02-exponential-smoothing-3 }

ts_marriages <- m |>
                mutate(date = yearmonth(date)) |>
                as_tsibble(index = date, key = c('code', 'name'))
glimpse(ts_marriages)

```

## ETS(ANN) model

**ETS** - **E**rror, **T**rend, **S**easonality.\
**ANN** - **A**dditive error, **N**o trend, **N**o seasonality.

-   $y_t$ - observed time series
-   $l_t$ - trend, cleaned time series
-   $u_t$ - random error

$$
y_t = l_{t-1} + u_t
$$

Assume

-   $l_t = l_{t-1} + \alpha u_t$, $l_0$ - start value.
-   $u_t \sim \mathcal{N}(0,\,\sigma^{2})$ and independent.

Parameters of the model: $\alpha, \sigma^2, l_0$.

$$
\begin{cases}
y_t = l_{t-1} + u_t \\
l_t = l_{t-1} + \alpha u_t, l_0 - \text {start value} \\
u_t \sim \mathcal{N}(0,\,\sigma^{2}) \text { and independent}
\end{cases} 
$$

### Forecast

#### One-step forecast {-}

$$
y_{T+1} = l_T + u_{T+1}
$$

$$
(y_{T+1}|\mathcal{F_T}) \sim \mathcal{N}(l_T, \sigma^2)
$$

#### Two-step forecast {-}

$$
y_{T+2} = l_{T+1} + u_{T+2} = l_T + \alpha u_{T+1} + u_{T+2}
$$

$$
(y_{T+2}|\mathcal{F_T}) \sim \mathcal{N}(l_T, \sigma^2(\alpha^2 + 1))
$$

#### Predictive Interval {-}

$$
[\hat l_T - 1.96\hat\sigma\sqrt{\hat\alpha^2+1}; \hat l_T + 1.96\hat\sigma\sqrt{\hat\alpha^2+1}]
$$


#### Simple exponential smoothing {-}

$$
\begin{cases}
y_t = l_{t-1} + u_t \\
l_t = l_{t-1} + \alpha u_t, l_0 - \text {start value}
\end{cases} 
$$

$$
l_t = l_{t-1} + \alpha(y_t-l_{t-1}) = \alpha y_t + (1-\alpha)l_{t-1}
$$

$$
\hat l_1 = y_1
$$

$$
\hat l_t = \alpha y_t + (1-\alpha)\hat l_{t-1}
$$

$$
\min_{\alpha} \sum (y_t - \hat l_t)^2
$$


### Example

```{r 02-exponential-smoothing-4 }

rf_m <- ts_marriages |>
  filter(code == 643, !is.na(total), yearmonth(date) >= yearmonth('2017 JAN'),
         yearmonth(date) <= yearmonth('2020 JAN'))
```

```{r 02-exponential-smoothing-5 }
# ETS(ANN)
rf_m |>
  filter(date < yearmonth('2019 Feb')) |>
  model(ETS(total ~ error('A') + trend('N') + season('N'))) |>
  forecast(rf_m |>
      filter(date >= yearmonth('2019 Feb')), h=12) |>
  autoplot(rf_m) +
  labs(title = 'ETS(ANN)')

```


## ETS(AAN) model


**ETS** - **E**rror, **T**rend, **S**easonality.\
**AAN** - **A**dditive error, **A**dditive trend, **N**o seasonality.

-   $y_t$ - observed time series
-   $l_t$ - trend, cleaned time series
-   $b_t$ - current speed of growth cleaned time series
-   $u_t$ - random error


$$
\begin{cases}
y_t = l_{t-1} + b_{t-1} + u_t \\
l_t = l_{t-1} + b_{t-1} + \alpha u_t, l_0 - \text {start value} \\
u_t \sim \mathcal{N}(0,\,\sigma^{2}) \text { and independent} \\
b_t = b_{t-1} + \beta u_t, \beta_0 - \text {start value}
\end{cases} 
$$


Parameters of the model: $\alpha, \beta, \sigma^2, l_0, b_0$.

### Forecast

#### One-step forecast {-}

$$
y_{T+1} = l_T + b_T + u_{T+1}
$$

$$
(y_{T+1}|\mathcal{F_T}) \sim \mathcal{N}(l_T + b_T, \sigma^2)
$$

#### Two-step forecast {-}

$$
y_{T+2} = l_{T+1} + b_{T+1} + u_{T+2} = (l_T + b_T+ \alpha u_{T+1}) + (b_T + \beta u_{T+1}) + u_{T+2}
$$

$$
(y_{T+2}|\mathcal{F_T}) \sim \mathcal{N}(l_T + 2b_T, \sigma^2((\alpha + \beta)^2 + 1))
$$

### Example

```{r 02-exponential-smoothing-6 }

# ETS(AAN)
rf_m |>
  filter(date < yearmonth('2019 Feb')) |>
  model(ETS(total ~ error('A') + trend('A') + season('N'))) |>
  forecast(rf_m |>
       filter(date >= yearmonth('2019 Feb')), h=12) |>
  autoplot(rf_m) +
  labs(title = 'ETS(AAN)')

```



## ETS(AAA) model

**ETS** - **E**rror, **T**rend, **S**easonality.\
**AAA** - **A**dditive error, **A**dditive trend, **A**dditive seasonality.

-   $y_t$ - observed time series
-   $l_t$ - trend, cleaned time series
-   $b_t$ - current speed of growth cleaned time series
-   $s_t$ - seasonal component
-   $u_t$ - random error


$$
\begin{cases}
y_t = l_{t-1} + b_{t-1} + s_{t-m} + u_t, m - \text{seasonal period} \\
l_t = l_{t-1} + b_{t-1} + \alpha u_t, l_0 - \text {start value} \\
u_t \sim \mathcal{N}(0,\,\sigma^{2}) \text { and independent} \\
b_t = b_{t-1} + \beta u_t, \beta_0 - \text {start value} \\
s_t = s_{t-m} + \gamma u_t, s_0, s_{-1},..., s_{-(m-1)} - \text{start values}
\end{cases} 
$$


Parameters of the model: $\alpha, \beta, \gamma, \sigma^2, l_0, b_0, s_0, s_{-1},...,s_{-(m-1)}$. \

Limitation: $s_0 + s_{-1} + ... + s_{-(m-1)} = 0$


### Forecast

#### One-step forecast {-}

$$
y_{T+1} = l_T + b_T + s_{T-11} + u_{T+1}
$$

$$
(y_{T+1}|\mathcal{F_T}) \sim \mathcal{N}(l_T + b_T + s_{T-11}, \sigma^2)
$$

#### Two-step forecast {-}

$$
y_{T+2} = l_{T+1} + b_{T+1} + s_{T-10} + u_{T+2} = (l_T + b_T+ \alpha u_{T+1}) + (b_T + \beta u_{T+1}) + s_{T-10} + u_{T+2}
$$

$$
(y_{T+2}|\mathcal{F_T}) \sim \mathcal{N}(l_T + 2b_T + s_{T-10}, \sigma^2((\alpha + \beta)^2 + 1))
$$

### Example

```{r 02-exponential-smoothing-7 }

# ETS(AAA)
rf_m |>
   filter(date < yearmonth('2019 Feb')) |>
  model(ETS(total ~ error('A') + trend('A') + season('A'))) |>
  forecast(rf_m |>
       filter(date >= yearmonth('2019 Feb')), h=12) |>
  autoplot(rf_m) +
  labs(title = 'ETS(AAA)')

```


## Model selection


### Mean Absolute Error {-}

$$
MAE = \frac {|e_{T+1}| + |e_{T+2}|+...+ |e_{T+H}|}{H}
$$


### Root Mean Squared Error {-}

$$
RMSE = \sqrt {\frac {e_{T+1}^2 + e_{T+2}^2+...+ e_{T+H}^2}{H}}
$$


### Mean Absolute Percentage Error {-}

$$
e_{t+h} = y_{t+h} - \hat y_{t+h}
$$

$$
p_t = e_t/y_t \times 100
$$

$$
MAPE = \frac {|p_{T+1}| + |p_{T+2}| + ... + |p_{T+H}|} H
$$


### Symmetric Mean Absolute Percentage Error {-}

$$
p_t^s = e_t/(0.5y_t + 0.5 \hat y_t) \times 100
$$


$$
sMAPE = \frac {|p_{T+1}^s| + |p_{T+2}^s| + ... + |p_{T+H}^s|} H
$$

### Mean Absolute Scaled Error {-}

$$
\hat y_t^{naive} = y_{t-1}
$$
$$
\hat y_t^{naive} = y_{t-12}
$$
$$
q_t = \frac {e_t}{MAE^{naive}}
$$

$$
MASE = \frac {|q_{T+1}| + |q_{T+2}|+...+|q_{T+H}|}{H}
$$


### Akaike Information Criterion {-}

$$
AIC = -2lnL + 3k
$$

## Examples

### Model selection by MSE

```{r 02-exponential-smoothing-8 }

glimpse(ts_marriages)

```

```{r 02-exponential-smoothing-9 }

rf_m <- ts_marriages |>
  filter(code == 643, !is.na(total))

rf_m |>
  gg_tsdisplay()

```


```{r 02-exponential-smoothing-10 }

train <- rf_m |>
  filter(date < yearmonth('2022 Jan'))

test <- rf_m |>
  anti_join(train, by = 'date')

```

```{r 02-exponential-smoothing-11 }

fits <- train |>
  model(snaive = SNAIVE(total),
        ets_aaa = ETS(total ~ error('A') + trend('A') + season('A')),
        ets_ana = ETS(total ~ error('A') + trend('N') + season('A')))

```

```{r 02-exponential-smoothing-12 }

fc <- fits |>
  forecast(h = '2 years')

```

```{r 02-exponential-smoothing-13 }

rf_m |>
  filter(date >= yearmonth('2015 JAN')) |>
  autoplot(total) +
  autolayer(fc |> filter(.model %in% c('snaive', 'ets_aaa'))) +
  facet_grid(.model ~ .)
  

```

```{r 02-exponential-smoothing-14 }

fc |>
  accuracy(test)

```


### Model Averaging

```{r 02-exponential-smoothing-15 }

fits <- train |>
  model(snaive = SNAIVE(total),
        ets_aaa = ETS(total ~ error('A') + trend('A') + season('A')),
        ets_ana = ETS(total ~ error('A') + trend('N') + season('A')),
        ets_aaa_ln = ETS(log(total) ~ error('A') + trend('A') + season('A')),
        ets_ana_ln = ETS(log(total) ~ error('A') + trend('N') + season('A')))

```


```{r 02-exponential-smoothing-16 }

fc <- fits |>
  forecast(h = '2 years')

```


```{r 02-exponential-smoothing-17 }

fc |>
  accuracy(rf_m)

```

```{r 02-exponential-smoothing-18 }

# make combinations
fits2 <- fits |>
  mutate(top2 = (ets_ana_ln + snaive)/2,
         top3 = (ets_ana_ln + snaive + ets_aaa_ln)/3)

```


```{r 02-exponential-smoothing-19 }

fits2 |>
  forecast(h = '2 years') |>
  accuracy(rf_m)

```


### Cross-validation

```{r 02-exponential-smoothing-20 }

rf_m |>
  nrow()

```

```{r 02-exponential-smoothing-21 }

rf_m_stretch <- rf_m |>
  stretch_tsibble(.init = 200, .step = 1)

```

```{r 02-exponential-smoothing-22 }

fits <- rf_m_stretch |>
  model(snaive = SNAIVE(total),
        ets_aaa = ETS(total ~ error('A') + trend('A') + season('A')),
        ets_ana = ETS(total ~ error('A') + trend('N') + season('A')),
        ets_aaa_ln = ETS(log(total) ~ error('A') + trend('A') + season('A')),
        ets_ana_ln = ETS(log(total) ~ error('A') + trend('N') + season('A')))

fits |>
  forecast(h = 1) |>
  accuracy(rf_m)

```

```{r 02-exponential-smoothing-23 }

fits2 <- fits |>
  mutate(top2 = (ets_ana_ln + ets_ana)/2)

fits2 |>
  forecast(h = 1) |>
  accuracy(rf_m)

```

