# ETS model variations

```{r 03-exponential-smoothing-1 }

# Libraries loading
library(tidyverse)
library(fpp3)
library(ggplot2)
library(rio)

```

```{r 03-exponential-smoothing-2 }

# data loading
m <- import(paste0(getwd(), '/00_data/marriages.csv'))
glimpse(m)

```

```{r 03-exponential-smoothing-3 }

ts_marriages <- m |>
                mutate(date = yearmonth(date)) |>
                as_tsibble(index = date, key = c('code', 'name'))
glimpse(ts_marriages)

```

```{r 03-exponential-smoothing-4 }

rf_m <- ts_marriages |>
  filter(code == 643, !is.na(total))

```

## ETS(AAdN): damped trend

In an **ETS model** with a **damped trend**, a special coefficient is
added to allow the trend's growth or decline to slow down over time.
This is useful when you expect that the trend will not continue to grow
or decrease at the same rate indefinitely but instead will gradually
slow down.

A **damping parameter** $\phi$ is introduced, which lies between 0
and 1. This parameter controls how the trend slows down. - If
$\phi = 1$, the trend remains additive (linear) and is not damped. - If
$\phi < 1$, the trend gradually reduces, slowing down over time.

In an **ETS(AAdN)** model (**A**dditive Error, **A**dditive **d**amped
Trend, **N**o Seasonality), the model is described as follows:

$$
\begin{cases}
y_t = \ell_{t-1} + \phi b _{t-1} + \beta u_t \\
b_t = \phi b _{t-1} + \beta u_t, b_0 - \text {start value} \\
l_t = \ell_{t-1} + \phi b _{t-1} + \alpha u_t, \ell_0 - \text {start value} \\
u_t \sim \mathcal{N}(0,\,\sigma^{2}) \text { and independent}
\end{cases}
$$

Parameters: $\alpha, \sigma^2, \ell_0, b_0, \beta, \phi$\

### Forecast {-}


#### One-step forecast {-}

$$
y_{T+1} = \ell_T + \phi b_T + u_{T+1}
$$

$$
(y_{T+1}|\mathcal{F_T}) \sim \mathcal{N}(l_T + \phi b_T, \sigma^2)
$$


#### Two-step forecast {-}

$$
y_{T+2} = l_{T+1} + \phi b_{T+1} + u_{T+2} = (l_T + \phi b_T + \alpha u_{T+1}) + \phi(\phi b_T + \beta u_{T+1}) + u_{T+2}
$$

$$
(y_{T+2}|\mathcal{F_T}) \sim \mathcal{N}(l_T + (\phi + \phi^2)b_T, \sigma^2((\alpha + \phi \beta)^2 + 1))
$$


### Example {-}

```{r 03-exponential-smoothing-5 }

rf_y <- rf_m |>
  index_by(year(date)) |>
  summarize(total = sum(total)) |>
  filter(`year(date)` < 2021)

rf_y|>
  model(ETS(total ~ error('A') + trend('Ad') + season('N'))) |>
  forecast(h = 10) |>
  autoplot(rf_y) +
  labs(title = 'ETS(AAdN): mariages in Russia', y = 'Mariages')

```

## ETS(MNM): multiplicative components

**ETS** - **E**rror, **T**rend, **S**easonality.\
**MNM** - **M**ultiplicative error, **N**o trend, **M**o seasonality.


$$
\begin{cases}
y_t = l_{t-1} \times s_{t-12} \times (1 + u_t) \\
l_t = l_{t-1} \times (1 + \alpha u_t), l_0 - \text {start value} \\
s_t = s_{t-12} \times (1 + \gamma u_t), s_0, ..., s_{-11} - \text {start values} \\
u_t \sim \mathcal{N}(0,\,\sigma^{2}) \text { and independent}
\end{cases} 
$$

-   $y_t$ - observed value at time $t$
-   $\ell_t$ - trend, or level at time $T$
-   $s_t$ - seasonal component
-   $u_t$ - random error

### Parameters {-}

-   Non-seasonal parameters: $\alpha, \sigma^2, l_0$
-   Seasonal parameters: $\gamma, s_0, s_{-1},..., s_{-11}$

Limitation: $s_0 \times s_{-1} \times...\times s_{-11} = 1$

-   $y_t, \ell_t$ - original units
-   $s_t, u_t$ - fraction
-   $s_t$ is measured relative to one, for example, $s_t = 0.9$ means it
    is 10% below the trend.
-   $u_t$ is measured relative to zero, for example, $u_t = -0.1$ means
    a 10% decrease.

### Forecast {-}

#### One-step forecast {-}

$$
y_{T+1} = l_T \times s_{T-11} \times (1 + u_{T+1})
$$

$$
(y_{T+1}|\mathcal{F_T}) \sim \mathcal{N}(\ell_T \times s_{T-11}, (\ell_T \times s_{T-11})^2\sigma^2)
$$

#### Two-step forecast {-}

$$
y_{T+2} = l_{T+1} \times s_{T-10} \times (1+u_{T+2}) = l_T (1+ \alpha u_{T+1}) \times s_{T-10} \times (1 + u_{T+2})
$$

### Example {-}

```{r 03-exponential-smoothing-6 }

fc1 <- aus_production |>
  filter(year(Quarter) < 1970) |>
  model(ETS(Electricity ~ error('M') + trend('N') + season('M'))) |>
  forecast(h = 16)

fc2 <- aus_production |>
  filter(year(Quarter) < 1980) |>
  model(ETS(Electricity ~ error('M') + trend('N') + season('M'))) |>
  forecast(h = 16)

aus_production |>
  filter(year(Quarter) < 1980) |>
  autoplot(Electricity) +
  autolayer(fc1) +
  autolayer(fc2) +
  labs(title = 'ETS(MNM): electricity production in Australia')

```


## ETS(MAdM): combinations

**MAdM**: **M**ultiplicative error, **A**dditive **d**umped trend,
**M**ultiplicative seasonality

$$
\begin{cases}
y_t = (l_{t-1} + \phi b_{t-1}) \times s_{t-12} \times (1 + u_t) \\
l_t = (l_{t-1} + \phi b_{t-1}) \times (1 + \alpha u_t), l_0 - \text {start value} \\
b_t = \phi b_{t-1} + \beta(l_{t-1} + \phi b_{t-1})u_t, b_0 - \text {start value}\\
s_t = s_{t-12} \times (1 + \gamma u_t), s_0, ..., s_{-11} - \text {start values} \\
u_t \sim \mathcal{N}(0,\,\sigma^{2}) \text { and independent}
\end{cases}
$$

### Parameters {-}

-   Non-seasonal parameters: $\alpha, \beta, \sigma, \phi, l_0, b_0$
-   Seasonal parameters: $\gamma, s_0, s_{-1}, ... , s_{-11}$

### Forecast {-}

#### One-step forecast {-}

$$
y_{T+1} = (l_T + \phi b_T) \times s_{T-11} \times (1 + u_{T+1})
$$

$$
(y_{T+1}|\mathcal{F_T}) \sim \mathcal{N}((\ell_T + \phi b_T) \times s_{T-11}, (\ell_T + \phi b_T)^2 \times s_{T-11}^2 \times \sigma^2)
$$

### Example {-}

```{r 03-exponential-smoothing-7 }

rf_m |>
  filter(year(date) > 2012, year(date) < 2022) |>
  model(ETS(total ~ error('M') + trend('Ad') + season('M'))) |>
  forecast(h = 48) |>
  autoplot(rf_m) +
  labs(title = 'EST(MAdM): forecast')

```


Combinations:

-   **Errors:** A, M
-   **Trend:** N, A, Ad, M, Md
-   **Seasonality:** N, A, M


Historic naming:

-   **ETS(ANN)** - Simple Exponential Smoothing
-   **ETS(AAA)** - Additive Holt-Winters Method
-   **ETS(AAM)** - Multiplicative Holt-Winters Method
-   **ETS(AAdM)** - Damped Additive Trend with Multiplicative Seasonality
    (Damped Holt-Winters)


## Examples

```{r 03-exponential-smoothing-8 }

rf_m |>
  gg_tsdisplay(total)

```

Split into training and test set.

```{r 03-exponential-smoothing-9 }

rf_train <- rf_m |>
      filter(year(date) < 2020)
rf_test <- rf_m |>
  anti_join(rf_train, by = 'date')

```


```{r 03-exponential-smoothing-10 }

models <- rf_train |>
  model(
    snaive = SNAIVE(total),
    ANA = ETS(total ~ error('A') + trend('N') + season('A')),
    AAA = ETS(total ~ error('A') + trend('A') + season('A')),
    AAdA = ETS(total ~ error('A') + trend('Ad') + season('A')),
    zzz = ETS(total), # select the best model by AIC
    azz = ETS(total ~ error('A')),
    theta = THETA(total) # theta method
  )

```


```{r 03-exponential-smoothing-11 }

models$zzz[[1]] |> report()

```

```{r 03-exponential-smoothing-12 }

models$theta[[1]] |> report()

```

```{r 03-exponential-smoothing-13 }

fc <- models |>
  forecast(h = 72)

```


```{r 03-exponential-smoothing-14 }

fc |> accuracy(rf_m)

```

```{r 03-exponential-smoothing-15, fig.width=6, fig.height=8}

rf_m |>
  autoplot(total) +
  autolayer(fc) +
  facet_grid(.model ~ .)

```
Composite model.

```{r 03-exponential-smoothing-16 }

models <- rf_train |>
  model(
    snaive = SNAIVE(total),
    theta = THETA(total),
    composit = decomposition_model(
      STL(total ~ season(window = Inf)),
      ETS(season_adjust ~ error('A') + trend('Ad') + season('N')),
      SNAIVE(season_year) ))

```


```{r 03-exponential-smoothing-17 }

models |>
  forecast(h = 48) |>
  accuracy(rf_m)

```

Model evaluation for multiple time series.

```{r 03-exponential-smoothing-18 }

train <- ts_marriages |>
  filter(year(date) < 2020)

```


```{r 03-exponential-smoothing-19 }
models <- train |>
  model(
    snaive = SNAIVE(total),
    zzz = ETS(log(total)), # select the best model by AIC
    theta = THETA(total)
  )

```

```{r 03-exponential-smoothing-20 }

acc <- models |>
  forecast(h = 48) |> 
  accuracy(rf_m)

```

```{r 03-exponential-smoothing-21 }

acc |>
  group_by(.model) |>
  summarise(avg_MAE = mean(MAE, na.rm = TRUE))

```

