# Stationary Process


```{r 04-stationarity-1 }

library(tidyverse)
library(fpp3)
library(ggplot2)
library(rio)

```

**Stationary process** is a random process with constant characteristics.

$$
\begin{cases}
\mathbb E(y_t) = \mu \\
Cov(y_t, y_{t+k}) = \gamma_k
\end{cases}
$$
+ Random walk is an example of non-stationary process
+ Random sample is an example of stationary process

## MA processes

For a process $y_t$ where $t \in \mathbb Z$, lagged process $Ly_t$ is the same sequence of values with shifted index:

$$
Ly_t = y_{t-1}
$$

$$
L^2y_t = L \cdot L \cdot y_t = L \cdot y_{t-1} = y_{t-2}
$$

$$
\bigtriangleup y_t = y_t - y_{t-1} = (1-L)y_t
$$

Seasonal difference:

$$
\bigtriangleup_{12} y_t = y_{t} - y_{t-12} = (1-L^{12})y_t
$$


A process $y_t$ that can be written as
$$
y_t = \mu + u_t +\alpha_1u_{t-1} +...+\alpha_qu_{t-q},
$$
where $\alpha_q \ne 0$ and $u_t$ is a white noise, we can name $MA(q)$ process.

> ACF of MA(q) process suddenly becomes zero for all lags $k > q$.

> PACF of MA(q) process convergates to zero exponentially fast.

> MA(q) process is stationary.


## Examples

```{r 04-stationarity-2 }

set.seed(777)

```


```{r 04-stationarity-3 }

data <- tibble(y = arima.sim(n = 120, model = list(ma = 0.9)))

data$date <- yearmonth(ymd('2000-12-01') + months(1:120))

data

```

```{r 04-stationarity-4 }

data <- as_tsibble(data, index = date)

data

```

```{r 04-stationarity-5 }

data |> gg_tsdisplay()

```

```{r 04-stationarity-6 }

data |>
  ACF() |>
  autoplot()

```

```{r 04-stationarity-7 }


data |>
  PACF() |>
  autoplot()


```

```{r 04-stationarity-8 }
m <- import(paste0(getwd(), '/00_data/marriages.csv'))
glimpse(m)
```

```{r 04-stationarity-9 }
ts_marriages <- m |>
                mutate(date = yearmonth(date)) |>
                as_tsibble(index = date, key = c('code', 'name'))
glimpse(ts_marriages)

```

```{r 04-stationarity-10 }

rf_m <- ts_marriages |>
  filter(code == 643, !is.na(total))

```


```{r 04-stationarity-11 }

rf_m |>
  gg_tsdisplay(total)

```

```{r 04-stationarity-12 }

rf_train <- rf_m |>
  filter(date < yearmonth('2022 May'))

mods <- rf_train |>
  model(snaive = SNAIVE(total),
         theta = THETA(total),
         ma12 = ARIMA(total ~ 1 + pdq(0, 0, 12) + PDQ(0, 0, 0)),
         stl_ma = decomposition_model(STL(total ~ season(window = Inf)),
                                      ARIMA(season_adjust ~ 1 + pdq(0, 0, 1:5) + PDQ(0, 0, 0)),
                                      SNAIVE(season_year)))

mods

```

```{r 04-stationarity-13 }

report(mods$ma12[[1]])

```

```{r 04-stationarity-14 }

report(mods$stl_ma[[1]])

```

```{r 04-stationarity-15 }

fc <- mods |>
  forecast(h = '2 years')

```


```{r 04-stationarity-16 }

fc |>
  accuracy(rf_m)

```



