# ARMA processes

```{r 05-ARMA-1 }

library(tidyverse)
library(fpp3)
library(ggplot2)
library(rio)

```

## AR process

An $\textbf{AR(p)}$ process is described by the following equation:
$$
y_t = c + \beta_1 y_{t-1} + \ldots + \beta_p y_{t-p} + u_t,
$$
where $u_t$ represents white noise, and $\beta_p \neq 0$. This process is called a solution of the $MA(\infty)$ relative to $u_t$.


$AR(p)$ process is __stationary__.

## ARMA process

An $\textbf{ARMA(p, q)}$ process is described by the following equation:

$$
y_t = c + \beta_1 y_{t-1} + \ldots + \beta_p y_{t-p} + u_t + \alpha_1 u_{t-1} + \ldots + \alpha_q u_{t-q},
$$

where $u_t$ represents white noise, $\beta_p \neq 0$, and $\alpha_q \neq 0$. This process is called a solution of the $MA(\infty)$ relative to $u_t$.


## Examples

```{r 05-ARMA-2 }

set.seed(777)

```


```{r 05-ARMA-3 }

# generate data
data <- tibble(a = arima.sim(n = 100,
                             model = list(ar = 0.5)),
               b = arima.sim(n = 100,
                             model = list(ar = 0.9)),
               c = cumsum(rnorm(n = 100, mean = 0, sd = 1)))

```


```{r 05-ARMA-4 }

data$year = 2001:2100
data <- data |> as_tsibble(index = year)

```


```{r 05-ARMA-5 }

data |>
  gg_tsdisplay(a, plot_type = 'partial')

```

```{r 05-ARMA-6 }

data |>
  gg_tsdisplay(b, plot_type = 'partial')

```

```{r 05-ARMA-7 }

data |>
  gg_tsdisplay(c, plot_type = 'partial')

```

```{r 05-ARMA-8 }

train <- data |> filter(year < 2081)

```


```{r 05-ARMA-9 }

mod_b <- train |>
  model(ar1 = ARIMA(b ~ pdq(1, 0, 0)), # is correct
        ma1 = ARIMA(b ~ pdq(0, 0, 1)),
        naive = NAIVE(b))

```


```{r 05-ARMA-10 }

mod_c <- train |>
  model(ar1 = ARIMA(c ~ pdq(1, 0, 0)),
        ma1 = ARIMA(c ~ pdq(0, 0, 1)),
        naive = NAIVE(c)) # is correct

```


```{r 05-ARMA-11 }

fc_b <- mod_b |> forecast(h = 20)
fc_c <- mod_c |> forecast(h = 20)

```


```{r 05-ARMA-12 }

fc_b |> autoplot(data |> filter(year > 2050))

```

```{r 05-ARMA-13 }

fc_c |> autoplot(data |> filter(year > 2050))

```

## Comparing AR, MA and ARMA

```{r 05-ARMA-14 }
m <- import(paste0(getwd(), '/00_data/marriages.csv'))
glimpse(m)
```

```{r 05-ARMA-15 }

m_agg <- m |> mutate(year = year(date)) |> 
  select(-date) |>
  group_by(code, name, year) |>
  summarise(sum = sum(total),
            max = max(total),
            .groups = 'keep')

glimpse(m_agg)

```

```{r 05-ARMA-16 }

rfy <- m_agg |> filter(code == 643) |>
  as_tsibble(index = year)

rfy |>
  gg_tsdisplay(sum, plot_type = 'partial')  

```

```{r 05-ARMA-17 }

train <- rfy |> filter(year < 2020)

models <- train |>
  model(ar1 = ARIMA(sum ~ pdq(1, 0, 0)),
        ma1 = ARIMA(sum ~ pdq(0, 0, 1)),
        naive = NAIVE(sum),
        arma11 = ARIMA(sum ~ pdq(1, 0, 1)))

```


```{r 05-ARMA-18 }

report(models$ar1[[1]])

```
```{r 05-ARMA-19 }

nrow(train)
nrow(rfy)

```

```{r 05-ARMA-20 }

fc <- models |>
  forecast(h = 5)

```


```{r 05-ARMA-21 }

fc |> accuracy(rfy)

```


