# ARIMA and seasonal ARIMA (sARIMA)

```{r 06-ARIMA-sARIMA-1 }

library(tidyverse)
library(fpp3)
library(ggplot2)
library(rio)
library(urca)

```

> How to choose $ARIMA(p,0,q), ARIMA(p,1,q)$ or $ARIMA(p,2,q)$?

1.  Looking at the plot. The variance of a stationary process is
    constant (approximately).
2.  Cross-validation
3.  Unit root test

-   ARMA: only for stationary process
-   ARIMA: for non-stationary process

## Unit root test

```{r 06-ARIMA-sARIMA-2 }
m <- import(paste0(getwd(), '/00_data/marriages.csv'))
glimpse(m)
```

```{r 06-ARIMA-sARIMA-3 }
ts_marriages <- m |>
                mutate(date = yearmonth(date)) |>
                as_tsibble(index = date, key = c('code', 'name'))
glimpse(ts_marriages)

```

```{r 06-ARIMA-sARIMA-4 }

rf_m <- ts_marriages |>
  filter(code == 643, !is.na(total))

```

```{r 06-ARIMA-sARIMA-5 }

rf_m |>
  gg_tsdisplay(total)

```

### KPSS test

```{r 06-ARIMA-sARIMA-6 }

kpss_res <- ur.kpss(rf_m$total, type = 'mu')

# H0: ts = mu + stat (stationary)
# Ha: ts = mu + stat + rw

summary(kpss_res)

# => H0 is rejected => difference is needed

```

### ADF test

```{r 06-ARIMA-sARIMA-7 }

adf_res <- ur.df(rf_m$total, type = 'drift', selectlags = 'AIC')

# H0: non-stationary process ts = ARIMA(p,1,q) + trend
# Ha: stationary process ts = ARIMA(p,0,q) + const

summary(adf_res)

# => H0 is rejected => difference is not needed

```

## Seasonal ARIMA

How to choose: $SARIMA(p, 0, q)(P, 0 , Q)$ or
$SARIMA(p, 0, q)(P, 1 , Q)[12]$?

### **Step 1**: How many times should I apply a seasonal differencing? {-}

-   **Step 1**: Apply STL decomposition
    $y_t = trend_t + season_t + remainder_t$
-   **Step 2**: Calculate $F_{season}$:
    $F_{season} = max\{1 - \frac {sVar(remainder)}{sVar(season + remainder)}, 0\}$
-   **Step 3**: If $F_{season}$ is below the threshold, then work with
    source series, otherwise, apply seasonal differencing. Repeat steps
    1-3 for seasonally differenced series and apply a second difference
    if it's needed.

### **Step 2**: How many times should I apply a first differencing? {-}

Apply the KPSS test with constant to the origin series:

-   If $H_0$ is not rejected, then work with the origin series,
    otherwise, apply the test to the first differenced series. Repeat
    the step, if $H_0$ is rejected, then work with the second
    differenced test, otherwise, work with the first differenced series.

### **Step 3**: Apply SARMA models to stationary series. {-}

-   Choose the best model by AIC (min.).

## Automatic ARIMA

```{r 06-ARIMA-sARIMA-8 }

# 1. Look at the plot, ACF, PACF

rf_m |>
  gg_tsdisplay(total, plot_type = 'partial')

```

Observations:

-   ACF is sinusoidal - it is typical for seasonal series
-   PACF: non-seasonal is exponentially decaying, but there is one
    significant seasonal lag. AR(1) p=1 for no-seasonal part, and AR(1)
    P=1 for seasonal part.

```{r 06-ARIMA-sARIMA-9 }

train <- rf_m |> filter(date <= yearmonth('2022 May'))

models <- train |>
  model(snaive = SNAIVE(total),
        theta = THETA(total),
        auto = ARIMA(total), # Khandakar-Hyndman Method
        sarima111_x11 = ARIMA(total ~ 0 + pdq(1,1,1) + PDQ(0:1,1,1))
  )

```

```{r 06-ARIMA-sARIMA-10 }

models$auto[[1]] |> report()

```

```{r 06-ARIMA-sARIMA-11 }

models$sarima111_x11[[1]] |> report()

```

```{r 06-ARIMA-sARIMA-12 }

fc <- models |>
  forecast(h = 2)

```

```{r 06-ARIMA-sARIMA-13 }

fc |> accuracy(rf_m) |> select(-code, -name)

```

Using cross validation.

```{r 06-ARIMA-sARIMA-14 }

m_slide = rf_m |> slide_tsibble(.size = 60, .step = 1)

m_slide

```

```{r 06-ARIMA-sARIMA-15 }

models_slide <- m_slide |>
  model(snaive = SNAIVE(total),
        theta = THETA(total),
        auto = ARIMA(total), # Khandakar-Hyndman Method
      )

```

```{r 06-ARIMA-sARIMA-16 }

fc_slide <- models_slide |> forecast(h = 1)

fc_slide |> accuracy(rf_m) |> select(-code, -name, -.type)

```

```{r 06-ARIMA-sARIMA-17 }

model_agg <- models_slide |> 
  mutate(av3 = (auto + snaive + theta)/3,
         auto_theta = (auto + theta)/2,
         snaive_theta = (snaive + theta)/2)

```

```{r 06-ARIMA-sARIMA-18 }

fc_slide_agg <- model_agg |> forecast(h = 1)

fc_slide_agg |> accuracy(rf_m) |> select(-code, -name, -.type)

```

## Output2Equations

```{r 06-ARIMA-sARIMA-19 }

train |>
  model(ARIMA(total ~ pdq(2,1,1) + PDQ(1,1,1))) |>
  report()

```

Interpretation:

-   ARIMA(2,1,1)(1,1,1)[12]: (2,1,1) - non-seasonal part, (1,1,1) -
    seasonal part:
    -   AR part:
        -   ARIMA(2,**1**,1)(1,1,1) - one non-seasonal lag, $d=1$:
            $(1 - L)$
        -   ARIMA(2,1,1)(1,**1**,1) - on seasonal lag, $D=1$:
            $(1 - L^{12})$
        -   ARIMA(**2**,1,1)(1,1,1) - AR non-seasonal part, $p=2$, two
            non-seasonal AR params: $ar1 = -0.9743$, $ar2 = -0.5254$:
            $(1 - (-0.9743) \cdot L - (-0.5254)\cdot L^2)$
        -   ARIMA(2,1,1)(**1**,1,1) - AR seasonal part, $p=1$, one
            seasonal AR param: $sar1 = 0.0343$:
            $(1 - (0.0343) \cdot L^{12})$
        -   **AR summary**:
            $(1 - L) \cdot (1 - L^{12}) \cdot (1 - (-0.9743) \cdot L - (-0.5254)\cdot L^2) \cdot (1 - (0.0343) \cdot L^{12}) \cdot y_t$
    -   MA part:
        -   ARIMA(2,1,**1**)(1,1,1) - one non-seasonal lag, $q=1$,
            $ma1 = 0.2502$: $(1 + 0.2502 \cdot L)$
        -   ARIMA(2,1,1)(1,1,**1**) - one seasonal lag, $Q=1$,
            $sma1 = -0.7840$: $(1 - 0.7840 \cdot L^{12})$
        -   **MA summary**:
            $(1 + 0.2502 \cdot L) \cdot (1 - 0.7840 \cdot L^{12}) \cdot u_t$,
            where $u_t$ is white noise.

The equation: $$
(1 - L) \cdot (1 - L^{12}) \cdot (1 - (-0.9743) \cdot L - (-0.5254)\cdot L^2) \cdot (1 - (0.0343) \cdot L^{12}) \cdot y_t = \\(1 + 0.2502 \cdot L) \cdot (1 - 0.7840 \cdot L^{12}) \cdot u_t
$$
