# ARIMA with predictors

```{r 07-ARIMA-predictors-1 }

library(tidyverse)
library(fpp3)
library(ggplot2)
library(rio)
library(urca)

```

## How to make predictors?

+ Using lags $y_{t-k}$
+ Using functions over lags

## Using lags

Training set:

$$
\begin{pmatrix}
  y_3 \\ 
  y_4 \\
  y_5 \\
  \vdots \\
  y_T
\end{pmatrix}
\begin{pmatrix}
  y_1 & y_2 \\
  y_2 & y_3 \\
  y_3 & y_4 \\
  \vdots & \vdots \\
  y_{T-2} & y_{T-1}
\end{pmatrix}
$$

> **Note!** Each added lag reduces the thaining set.


## Using function of lags

+ __Aggregating functions__: minimum, maximum, mean, median, range, variance, standard deviation,...
+ An argument of aggregating function:
  + Moving window
  + Growing window
  

Example of maximum over a moving window and minimum over a growing window:

$$
\begin{pmatrix}
  y_3 \\ 
  y_4 \\
  y_5 \\
  \vdots \\
  y_T
\end{pmatrix}
\begin{pmatrix}
  max(y_1, y_2) & min(y_1, y_2) \\
  max(y_2, y_3) & min(y_1, y_2, y_3) \\
  max(y_3, y_4) & min(y_1, \dots , y_4) \\
  \vdots & \vdots \\
  max(y_{T-2}, y_{T-1}) & min(y_1,\dots, y_{T-1})
\end{pmatrix}
$$

## Time predictors

+ Trend predictors
+ Seasonal and holidays dummies
+ Sin and Cos over time


## Trend predictors


$$
\begin{pmatrix}
  y_3 \\ 
  y_4 \\
  y_5 \\
  \vdots \\
  y_T
\end{pmatrix}
\begin{pmatrix}
  1 & \sqrt 1 \\
  2 & \sqrt 2 \\
  3 & \sqrt 3 \\
  \vdots & \vdots \\
  T & \sqrt T
\end{pmatrix}
$$

Forecast:

$$
\begin{pmatrix}
  ?
\end{pmatrix}
\begin{pmatrix}
  T+1 & \sqrt {T+1}
\end{pmatrix}
$$

A monotone transformation:

+ It is __useless__ for tree-based algorithms
+ The origin variable could be transformed: log, Box-Cox,...

## Seasonal and holidays dummies

Seasonal dummies (for quarterly seasonality):

$$
\begin{pmatrix}
  y_1 \\
  y_2 \\
  y_3 \\ 
  y_4 \\
  y_5 \\
  y_6 \\
  \vdots \\
  y_T
\end{pmatrix}
\begin{pmatrix}
  1 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 \\
  0 & 0 & 1 & 0 \\
  0 & 0 & 0 & 1 \\
  1 & 0 & 0 & 0 \\
  0 & 1 & 0 & 0 \\
  \vdots & \vdots & \vdots & \vdots \\
  0 & 0 & 1 & 0 \\
\end{pmatrix}
$$
> __Important!__ For multiple regression with seasonal dummy variables, you have two options: either include a dummy variable for each season and omit the constant (intercept), or include the intercept and use dummy variables for all but one season.

## Sin and Cos over time

For daily data:

+ Add predictor $cos(\frac {2\pi} {365} \cdot t)$ and $sin(\frac {2\pi} {365} \cdot t)$
+ Add predictor $cos(\frac {2\pi} {365} \cdot 2t)$ and $sin(\frac {2\pi} {365} \cdot 2t)$
+ Add predictor $cos(\frac {2\pi} {365} \cdot 3t)$ and $sin(\frac {2\pi} {365} \cdot 3t)$
+ $\dots$

## ARIMA and predictors

+ Linear regression with ARMA errors
+ ARMAX model
+ ARDL model

### Example 1

```{r 07-ARIMA-predictors-2 }

library(caret)
library(ranger)
library(xgboost)

```


```{r 07-ARIMA-predictors-3 }

air <- as_tsibble(AirPassengers)

gg_tsdisplay(air)

```

Make a variance stable.

```{r 07-ARIMA-predictors-4 }

air2 <- air |>
  mutate(ln_pass=log(value), t = 1:nrow(air))

glimpse(air2)

```
```{r 07-ARIMA-predictors-5 }

air2 |>
  gg_tsdisplay(ln_pass)

```

```{r 07-ARIMA-predictors-6 }

fourier_x <- forecast::fourier(AirPassengers, K = 2)

fourier_x

# sin(2pi t/12) cos(2pi t/12)
# sin(2pi t/12 * 2) cos(2pi t/12 * 2)

```

```{r 07-ARIMA-predictors-7 }

colnames(fourier_x) <- c('s1', 'c1', 's2', 'c2')

```

```{r 07-ARIMA-predictors-8 }

air3 <- bind_cols(air2, fourier_x)

glimpse(air3)

```

```{r 07-ARIMA-predictors-9 }

air3_test <- tail(air3, 24)
air3_train <- head(air3, -24)

autoplot(air3_train, ln_pass) +
  autolayer(air3_test, ln_pass, col='red')

```


Linear regression:

```{r 07-ARIMA-predictors-10 }

cv_params <- trainControl(method = 'cv', number = 5)

ols <- train(ln_pass ~ t + s1 + c1 + s2 + c2,
             data = air3_train,
             trControl = cv_params,
             method = 'lm')

```

Random Forest:

```{r 07-ARIMA-predictors-11 }

rf <- train(ln_pass ~ t + s1 + c1 + s2 + c2,
             data = air3_train,
             trControl = cv_params,
             method = 'ranger', num.trees = 10000)

```

Gradient Boosting:

```{r 07-ARIMA-predictors-12, warning=FALSE}

gb <- train(ln_pass ~ t + s1 + c1 + s2 + c2,
             data = air3_train,
             trControl = cv_params,
             method = 'xgbTree')

```

```{r 07-ARIMA-predictors-13 }

ols_fcst <- predict(ols, air3_test)
rf_fcst <- predict(rf, air3_test)
gb_fcst <- predict(gb, air3_test)

```


```{r 07-ARIMA-predictors-14 }

air4_test <- air3_test |>
  mutate(ols = ols_fcst,
         rf = rf_fcst,
         gb = gb_fcst)

```


```{r 07-ARIMA-predictors-15 }

air4_test |>
  glimpse()

```

```{r 07-ARIMA-predictors-16 }

air3_train |>
  autoplot(ln_pass) +
  autolayer(air4_test, ln_pass) +
  autolayer(air4_test, ols, col='red') +
  autolayer(air4_test, rf, col='green') +
  autolayer(air4_test, gb, col='blue')

```

### Example 2. Daily data

```{r 07-ARIMA-predictors-17 }

library(tsibbledata)

```


```{r 07-ARIMA-predictors-18 }

head(vic_elec)

```

```{r 07-ARIMA-predictors-19 }

elec <- vic_elec |>
  index_by(Date) |>
  summarise(dem = mean(Demand), temp = mean(Temperature))

gg_tsdisplay(elec)

```

```{r 07-ARIMA-predictors-20 }

gg_tsdisplay(tail(elec, 60))

```

```{r 07-ARIMA-predictors-21 }

elec_train <- head(elec, -60)
elec_test <- tail(elec, 60)

```


```{r 07-ARIMA-predictors-22 }

mods <- elec_train |>
  model(
    naive = NAIVE(dem),
    arima_fK1 = ARIMA(dem ~ fourier(K = 1) + PDQ(0, 0, 0)),
    arima_fK2 = ARIMA(dem ~ fourier(K = 2) + PDQ(0, 0, 0)),
    arima_fK3 = ARIMA(dem ~ fourier(K = 3) + PDQ(0, 0, 0))
  )

```

```{r 07-ARIMA-predictors-23 }

frct <- mods |>
  forecast(h = 60)

frct |>
  autoplot(tail(elec_train, 30)) +
  facet_grid(.model ~ .)

```

```{r 07-ARIMA-predictors-24 }

frct |> accuracy(elec)

```


### Example 3

```{r 07-ARIMA-predictors-25 }

library(ARDL)

```

```{r 07-ARIMA-predictors-26 }

glimpse(denmark)

```

```{r 07-ARIMA-predictors-27 }

time(denmark)

```

```{r 07-ARIMA-predictors-28 }

den <- as_tibble(denmark)

den

```

```{r 07-ARIMA-predictors-29 }

den2 <- den |> mutate(quarter = yearquarter(time(denmark)))
den2

```

```{r 07-ARIMA-predictors-30 }

den3 <- as_tsibble(den2, index = quarter)

den3 |>
  gg_tsdisplay(LRM)

```

```{r 07-ARIMA-predictors-31 }

ardl1 <- ardl(data = denmark, LRM ~ LRY + IBO + IDE,
              order = c(2, 1, 2, 2))

ardl1

```

```{r 07-ARIMA-predictors-32 }

ardl_many <- auto_ardl(data = denmark, LRM ~ LRY + IBO + IDE, max_order = 3)

ardl_many

```

```{r 07-ARIMA-predictors-33 }

ardl_many$best_model

```

