# Missing values, anomalies, structural shift


```{r 08-missing-1 }

library(tidyverse)
library(fpp3)
library(ggplot2)
library(rio)
library(urca)
library(imputeTS)

```

## Missing values

Data loading.

```{r 08-missing-2 }

air <- AirPassengers |>
  as_tsibble()

air |>
  gg_tsdisplay()

```

Create some NA's.

```{r 08-missing-3 }

lair <- log(AirPassengers)
lair_na <- lair
where_na <- c(5:6, 30:32, 70, 90:91, 110, 124)

lair_na[where_na] <- NA

```

Plot NA's.

```{r 08-missing-4 }

ggplot_na_distribution(lair_na)

```

```{r 08-missing-5 }

ggplot_na_distribution2(lair_na)

```

### Imputation with linear interpolation

```{r 08-missing-6 }

# linear interpolation

lair_int <- na_interpolation(lair_na)

ggplot_na_imputations(lair_na, lair_int, lair)

```

### ARIMA imputation

```{r 08-missing-7 }

# ARIMA imputation

mod <- arima(lair_na, order = c(1,1,0), seasonal = c(0,1,0))$model

lair_arima <- na_kalman(lair_na, model = mod)
lair_arima.auto <- na_kalman(lair_na, model = 'auto.arima')

ggplot_na_imputations(lair_na, lair_arima.auto, lair)

```

### Imputation with STL decomposition

```{r 08-missing-8 }

# STL decomposition

lair_seas <- na_seadec(lair_na)
ggplot_na_imputations(lair_na, lair_seas, lair)

```

## Outliers and anomalies

```{r 08-missing-9 }

library(anomalize)
library(tibbletime)
library(lubridate)

```

Make a table from `AirPAssenger`.

```{r 08-missing-10 }

air <- tibble(pass=AirPassengers)

head(air)

```

Add a new column for date.

```{r 08-missing-11 }

n <- nrow(air)
air2 <- air |> mutate(date = ymd('1949-01-01') + months(0:(n-1)))

air2

```
Plot the data.

```{r 08-missing-12 }

qplot(data=air2, x = date, y = pass, geom = 'line')

```

Make a log-transformation to stabilize the variance of the data.

```{r 08-missing-13 }

air3 <- air2 |> mutate(ln_pass = log(pass))
air3

```

Plot the transformed data.

```{r 08-missing-14 }

qplot(data=air3, x = date, y = ln_pass, geom = 'line')

```

Make a couple of anomalies to try to identify them.

```{r 08-missing-15 }

air4 <- air3 |> mutate(ln_pass_an = ln_pass)
air4$ln_pass_an[30] = 6
air4$ln_pass_an[110] = 5.3

qplot(data=air4, x = date, y = ln_pass_an, geom = 'line')

```

Run a STL decomposition to identify anomalies.

```{r 08-missing-16 }

decomp <- time_decompose(air4, target = ln_pass_an, frequency = 12)

decomp

```

Identify anomalies by reminder component.

```{r 08-missing-17 }

decomp2 <- anomalize(decomp, target = remainder)

glimpse(decomp2)

```

**Columns:**

+ **date:** date of observation
+ **observed:** observed value
+ **season:** seasonal component
+ **trend:** trend component
+ **remainder:** remainder component
+ **remainder_l1:** left bound of the normal value
+ **remainder_l2:** right bound of the normal value
+ **anomaly:** whether the observation is an anomaly or not


Plot the anomalies.

```{r 08-missing-18 }

decomp2 |> plot_anomalies()

```

Recompose the left and right bounds into the original variable (transformed).

```{r 08-missing-19 }

decomp3 <- time_recompose(decomp2)
glimpse(decomp3)

```

Plot the decomposition.

```{r 08-missing-20 }

plot_anomaly_decomposition(decomp2)

```

Clean the data into the origin series (transformed).

```{r 08-missing-21 }

decomp4 <- clean_anomalies(decomp3)
glimpse(decomp4)

```
The data are now without anomalies.

```{r 08-missing-22 }

qplot(data=decomp4, x = date, y = observed_cleaned, geom = 'line')

```

## Structural shift

```{r 08-missing-23 }

library(changepoint)

```



```{r 08-missing-24 }

m <- import('https://github.com/akarlinsky/world_mortality/raw/main/world_mortality.csv')
glimpse(m)

```
```{r 08-missing-25 }

m2 <- m |>
  filter(time_unit == 'monthly', country_name == 'Russia') |>
  mutate(date = ymd(paste0(year, '-', time, '-01'))) |>
  select(date, deaths)

qplot(data = m2, x = date, y = deaths, geom = 'line')

```

Make a STL decomposition.

```{r 08-missing-26 }

nm3 <- m2 |>
  as_tbl_time(index = date)

decomp <- time_decompose(nm3, target = deaths)

```

```{r 08-missing-27 }

decomp

```

Find a structural shift.

```{r 08-missing-28 }

# AMOC = At Most One Changepoint
one_break <- cpt.mean(nm3$deaths, method = 'AMOC')

one_break

```

Look at the structural shift.

```{r 08-missing-29 }

nm3[64, ]

```
Plot the structural shift.

```{r 08-missing-30 }

plot(one_break)

# => Assumption: The structural shift because of the COVID

```

Apply 3 structural shifts searching.

```{r 08-missing-31 }

# BigSeg = Binary Segmentation
all_breaks <- cpt.mean(nm3$deaths, method = 'BinSeg', Q = 3)

all_breaks
plot(all_breaks)

```

Apply to decomposed series.

```{r 08-missing-32 }

# AMOC = At Most One Changepoint 
one_break <- cpt.mean(decomp$remainder, method = 'AMOC')
one_break
plot(one_break)

```

## Bayesian Structural Model 

```{r 08-missing-33 }

library(bsts)

```

Plot the data.

```{r 08-missing-34 }

plot(AirPassengers)

```

Log the data to stabilize the variance.

```{r 08-missing-35 }

log_air <- log(AirPassengers)
plot(log_air)

```

Observations:

+ Stable variance of the transformed data
+ There is a linear trend
+ There is a seasonality


**Step by step:** \


1. Create an empty list

```{r 08-missing-36 }

model = list()

```

2. Add a trend

```{r 08-missing-37 }

model <- AddLocalLinearTrend(model, y = log_air)

```

3. Add a season

```{r 08-missing-38 }

model <- AddTrig(model, y = log_air, period = 12, frequencies = 1:2)

```

4. Run generation

```{r 08-missing-39 }

poster <- bsts(log_air, state.specification = model, niter = 2000)

```
5. Plot the data

```{r 08-missing-40 }

plot(poster, 'components')

```

6. Make forecasts.

```{r 08-missing-41 }

frst <- predict(poster, horizon = 24, quantiles = c(0.05, 0.95),
                burn = 1000)

frst$mean

```

```{r 08-missing-42 }

plot(frst)

```

## Causal impact

```{r 08-missing-43 }

library(CausalImpact)

```

```{r 08-missing-44 }

m <- import('https://github.com/akarlinsky/world_mortality/raw/main/world_mortality.csv')
glimpse(m)

```
```{r 08-missing-45 }

m2 <- m |>
  filter(time_unit == 'monthly', country_name == 'Russia') |>
  mutate(date = ymd(paste0(year, '-', time, '-01'))) |>
  select(date, deaths)

qplot(data = m2, x = date, y = deaths, geom = 'line')

```

```{r 08-missing-46 }

start_covid <- 61 # number of observation when the Covid has been started

impact <- CausalImpact(data = m2$deaths, pre.period = c(1, start_covid-1), 
                       post.period = c(start_covid, nrow(m2)))

```


```{r 08-missing-47}

impact

```


```{r 08-missing-48 }

plot(impact)

```


```{r 08-missing-49 }

summary(impact)

```


```{r 08-missing-50 }

summary(impact, 'report')

```




